Translation of B Implementations to the LLVM: Control and Modularity

### 2013-08-29 translation.tex
Restored specification of the translation for the CASE instruction. Created a B implementation with a CASE statement. Created a Python data structure representing that B implementation. Updated the Python module bimp to handle the new constructs (case, multiplication). Created a LLVM scaffolding to instantiate the B implementation and call its operations. Implemented the translation of the CASE statement to LLVM-IR.

        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \\

@@

%         \\
%         \hline
%         \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
%         \hline
%         \B{expr} & : & \B{Expr} \\
%         \B{branches} & : & seq \B{CaseBr}

@@

%         \hline
%         \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
%         \hline
%         \B{val} & : & seq \B{Value} \\
%         \B{body} & : & \B{Inst} \\

@@

@@

@@

>>>>

        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case}} \alt \B{Skip} \\

@@

        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
        \B{expr} & : & \B{Expr} \\
        \B{branches} & : & seq \B{CaseBr}
        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
        \B{val} & : & seq \B{Value} \\

@@

@@

\DD{Specify \B{Skip}.}

@@

\paragraph{Case instructions} are formed by an expression, and a list of case
branches, each of which has a list of values and an instruction block.
Optionally, there might be a default branch, and we consider that, in this case,
the list of values is empty.

The translation to the LLVM IR uses its switch statement. It is composed of a
variable, a default jump and a jump table. The jump table is a sequence of pairs
formed by a value, and an instruction label. In the following, \llvm{\%val} is
the variable, \llvm{\%otherwise} is the default label, and the jump table has
three entries. The first corresponds to the case where the value of \llvm{\%val}
is \llvm{0}, in which case the execution jumps to the label \llvm{onzero}.
\begin{verbatim}
switch i32 %val, label %otherwise [ i32 0, label %onzero
                                    i32 1, label %onone
                                    i32 2, label %ontwo ]
\end{verbatim}

The translation of the case statement first consists in producing the code $p$
to evaluate the expression tested in the statement. When that code completes
execution, the value is stored in variable $v$ of type $t$. Then the jump table
$j$ and the code $b$ of the different branches are generated. This generation
takes also as parameters labels $\ell$ and $\ell_o$.
\begin{align*}
  \trad{v, $\ell$}{\B{Case}} \isdef & \LET p \sep v \sep t = \trad{v.expr}{\B{Expr}} \IN \\
  & \LET \ell_o = \lbl \IN \\
  & \LET j, b = \trad{v.branches, $\ell_o$, $\ell$}{\ListOf{CaseBr}} \IN \\
  & \quad \llvm{"\PH{p}} \\
  & \quad \llvm{switch \PH{t} \PH{v}, label \%\PH{\ell_o} [ \PH{j} ] \PH{b}"}
\end{align*}

The following rule defines the composition of the translation of the different
branches of a case statement. Again, this translation has two parts: bits of the
jump table $j$, and instruction blocks $b$. If a case branch has several values,
then several entries need to be created in the jump table. This is the role
of the translation function $\trad{}{\B{ListOf{CaseVal}}}$.
\begin{align*}
  \trad{b bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \isdef 
  & \LET \ell = \lbl \IN \\
  & \LET j, b = \trad{bl, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \IN \\
  & \quad \llvm{"\PH{\trad{b.values, $\ell$}{\ListOf{\B{CaseVal}}}} \PH{j}"}
  \sep \\
  & \quad
  \llvm{"\PH{\ell}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L} \PH{b}"}
\end{align*}
This rule handles the last branch of a case statement, and needs to test if
that branch is a default branch or not. If not, then one needs to be artificially
created (as this is mandatory in the LLVM-IR switch statement):
\begin{align*}
  \trad{b, $\ell_o$, $\ell_e$}{\ListOf{\B{CaseBr}}} \isdef 
  & \IF \B{b.values} = \text{None} \THEN \\
  & \quad \llvm{""} \sep \llvm{"\PH{\ell_o}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L}"} \\
  & \ELSE \LET \ell_b = \lbl \IN \\
  & \quad \trad{b.values, $\ell_b$}{\ListOf{\B{CaseVal}}} \sep \\
  & \quad \llvm{"\PH{\ell_b}: \nl \PH{\trad{b.body, $\ell_e$}{\ListOf{Inst}}_L}} \\
  & \quad \quad \llvm{\PH{\ell_o}: \nl br label \% \PH{\ell_e}}
\end{align*}

The following rules specifies that the values of a case branch are translated in
order, their translations being concatenated to form the result. The parameter
$\ell$ is the label of the instruction block containing the translation of the
branch body:
\begin{align*}
  \trad{v, $\ell$}{\ListOf{\B{CaseVal}}} \isdef & \trad{v, $\ell$}{\B{CaseVal}} \\
  \trad{v vl, $\ell$}{\ListOf{\B{CaseVal}}} \isdef & 
  \llvm{"\PH{\trad{v, $\ell$}{\B{CaseVal}}} \PH{\trad{vl, $\ell$}{\ListOf{\B{CaseVal}}}}"}
\end{align*}

This rule defines how a single value of a case branch is translated, namely
as an entry in the jump table of a switch statement:
\begin{align*}
  \trad{v, $\ell$}{\B{CaseVal}} \isdef & \LET p \sep v \sep t = \trad{v}{\B{Expr}} \IN \\
  & \quad \llvm{"\PH{t} \PH{v}, label \%\PH{\ell} \nl"}
\end{align*}

@@

\newpage
\appendix

\section{Test summary}

For each test file, indicates the B0 constructs found in the implementation.
\begin{center}
\newcommand{\no}[0]{--}
\newcommand{\yes}[0]{$\circ$}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Example nº & 01 & 02 & 03 & 04 & 05 & 06 \\
\hline
variables & \yes & \yes & \yes & \yes & \yes & \yes\\
constants & \no & \yes & \no & \no & \yes & \no \\
op. input & \yes & \no & \yes & \no & \no & \no \\
op. output & \yes & \no & \yes & \no & \no & \no \\
\hline
\multicolumn{7}{|c|}{instructions} \\
\hline
block & \yes & \no & \yes & \yes & \no & \yes\\
local var & \no & \no & \yes & \yes & \no & \yes \\
identity & \no & \no & \yes & \no & \no & \no \\
becomes equal & \yes & \yes & \yes & \yes & \yes & \yes\\
operation call & \no & \no & \no & \no & \no & \no \\
condition & \yes & \no & \no & \no & \yes & \no \\
case & \no & \no & \no & \no & \no & \yes \\
assertion & \no & \no & \no & \no & \no & \no \\
sequence & \yes & \yes & \yes & \yes & \no & \yes \\
while & \no & \no & \no & \yes & \no & \no\\
\hline
\multicolumn{7}{|c|}{expressions} \\
\hline
int lit & \yes & \yes & \yes & \yes & \yes & \yes\\
bool lit & \yes & \yes & \no & \yes & \yes & \no \\
vars & \yes & \yes & \yes & \yes & \no & \yes \\
loc. vars & \no & \no & \yes & \yes & \no & \yes \\
const & \no & \yes & \no & \no & \yes & \no \\ 
succ & \no & \yes & \no & \yes & \yes & \no \\
pred & \no & \yes & \no & \yes & \yes & \no \\
+ & \yes & \no & \no & \yes & \no & \yes \\
- & \yes & \no & \no & \no & \no & \no \\
* & \no & \no & \no & \no & \no & \yes \\
\hline
\multicolumn{7}{|c|}{formulas} \\
\hline
and & \no & \no & \no & \no & \yes & \no \\
or & \no & \no & \no & \no & \yes & \no \\
not & \no & \no & \no & \no & \yes & \no \\
$<$ & \yes & \yes & \no & \no & \no & \no \\
$>$ & \no & \no & \no & \no & \yes & \no \\
\hline
\end{tabular}
\end{center}
